Explore consensus cell types

### 2025-01-31 analyses/cell-type-consensus/exploratory-notebooks/02-explore-consensus-results.Rmd
explore pdx vs patient tissue in consensus results

  summary_df <- df |> 
    dplyr::group_by(library_id, consensus_annotation, consensus_ontology) |> 
    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
    dplyr::left_join(total_cells_df, by = "library_id") |> 
    dplyr::mutate(
      # add percentage 
      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
    ) |> 
    dplyr::ungroup()

@@

@@

    # make a stacked bar chart with top cell types 
    ggplot(project_df) + 
      aes(
        x = library_id, 
        y = percent_cells_annotation, 
        fill = top_celltypes
      ) +
      geom_col() + 
      scale_y_continuous(expand = c(0,0)) +
      scale_fill_manual(values = colors, name = "cell type") +
      ggtitle(label) +
      theme(axis.text.x = element_blank())

@@

- Cell types identified tend to line up with expectations for the type of tumor. 
For example, leukemia libraries have T and B cells, brain tumors have macrophages, and solid tumors have fibroblasts and muscle cells. 
- Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
Notably many of the solid tumor projects (4, 5, 12-16, and 23) have a handful of PDX samples where I would expect to see fewer normal cells.

>>>>

  summary_df <- df |> 
    dplyr::group_by(library_id, sample_type, consensus_annotation, consensus_ontology) |> 
    dplyr::summarize(total_cells_per_annotation = length(consensus_annotation)) |>
    dplyr::left_join(total_cells_df, by = "library_id") |> 
    dplyr::mutate(
      # add percentage 
      percent_cells_annotation = round((total_cells_per_annotation / total_cells_per_library) * 100, 2)
    ) |> 
    dplyr::ungroup()

@@

Let's also look at the projects that have PDX samples. 
I expect that fewer cells are able to be classified in PDX samples compared to their patient tissue counterparts.

```{r, fig.height=7}
# list of projects with pdx 
pdx_projects <- c(
  "SCPCP000003",
  "SCPCP000004",
  "SCPCP000005",
  "SCPCP000011",
  "SCPCP000013",
  "SCPCP000014",
  "SCPCP000015",
  "SCPCP000016",
  "SCPCP000023"
)

# filter to pdx projects and adjust the facet label 
pdx_only <- unknown_only |>
  dplyr::filter(project_id %in% pdx_projects) |>
  dplyr::mutate(facet_label = glue::glue("{project_id}\n{diagnosis}"))

# faceted sina showing patient tissue vs. pdx 
ggplot(pdx_only, aes(x = sample_type, y = percent_cells_annotation)) +
  ggforce::geom_sina() +
  stat_summary(fun.y=median, geom="crossbar" , color = "red", linewidth = 0.2) +
  facet_wrap(vars(facet_label)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        plot.margin = margin(10,10,10,10)) +
  labs(
    x = "", 
    y = "Percent of cells annotated as Unknown"
  )
```

It looks like in `SCPCP000003` and `SCPCP000004` the PDX samples have less cells classified than the patient tissue samples.

For `SCPCP000015` there may be a higher percentage of cells that are classified in a few of the PDX samples. 
For `SCPCP000023` and `SCPCP000013` there appears to be a handful of PDX and patient tissue samples that have cells that are classified with no real difference between the two sample types. 
My hunch is that the cells getting classified in `SCPCP000013`, `SCPCP000015`, and `SCPCP000023` PDX samples are muscle cells which are very similar to the tumor cells.

For the remaining projects, there are very few samples that have any cells classified regardless of sample type.

@@

    # make a stacked bar chart with top cell types 
    ggplot(project_df) + 
      aes(
        x = library_id, 
        y = percent_cells_annotation, 
        fill = top_celltypes
      ) +
      geom_col() + 
      # split samples based on sample type, patient tissue or pdx 
      facet_wrap(vars(sample_type), scales ="free") +
      scale_y_continuous(expand = c(0,0)) +
      scale_fill_manual(values = colors, name = "cell type") +
      ggtitle(label) +
      theme(axis.text.x = element_blank())

@@

- Cell types identified tend to line up with expectations for the type of tumor. 
For example, leukemia libraries have T and B cells, brain tumors have macrophages, and solid tumors have fibroblasts and muscle cells. 
- Projects that I would expect to be more difficult to classify (sarcomas, wilms, RB) have fewer cells classified then things like brain and leukemia. 
- Generally there are very few cells classified in the PDX samples, and if they are classified they tend to be muscle cell types, fibroblasts, and in a handful of cases, endothelial cells. 
Immune cell types appear to be absent. 
Notably the projects that have PDX samples are solid tumors and sarcoma projects, where muscle cells or fibroblasts can be similar to tumor cells. 
- `SCPCP000023` seems to be an exception where a good amount of PDX samples have a high percentage of classified cells, although they are either muscle cells or chondrocytes. 
My guess is that these cells look very similar to tumor cells, which is why they are being classified.

It's important to note that when we quantify gene expression using `alevin-fry` in `scpca-nf` we only use a human reference, so I would not expect non-malignant cells to appear in the PDX samples at any high frequency. 
Because of this, I think it may be safest to remove these samples when conducting any further analysis.
